---
title: "analysis_1"
---
```{r}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, ggplot2, dplyr, lubridate, stringr, readxl, data.table, gdata,modelsummary,rdrobust,readr,lmtest,sandwich,broom,knitr,fixest,scales)

acs_medicaid <- read_tsv('/Users/sushmitarajan/econ470spring2025/Homework5/data/output/acs_medicaid.txt')

```

1. Plot the share of the adult population with direct purchase health insurance over time.

```{r}
# Calculate the share by year
direct_share_by_year <- acs_medicaid %>%
  group_by(year) %>%
  summarise(
    total_direct = sum(ins_direct, na.rm = TRUE),
    total_adults = sum(adult_pop, na.rm = TRUE),
    direct_share = total_direct / total_adults
  )

# Plot
q1<- ggplot(direct_share_by_year, aes(x = year, y = direct_share)) +
  geom_line(color = "black", size = 1.2) +
  geom_point(color = "black", size = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  labs(
    title = "Share of Adult Population with Direct Purchase Health Insurance",
    x = "Year",
    y = "Direct Purchase Share (%)"
  ) +
  theme_minimal()

q1
```

2. Discuss the reduction in direct purchase health insurance in later years. Can you list a couple of policies that might have affected the success of the direct purchase insurance market?

```{r}






```

3. Plot the share of the adult population with Medicaid over time.

```{r}
# Summarize by year: total Medicaid-insured and adult population
medicaid_share_by_year <- acs_medicaid %>%
  group_by(year) %>%
  summarise(
    total_medicaid = sum(ins_medicaid, na.rm = TRUE),
    total_adults = sum(adult_pop, na.rm = TRUE),
    medicaid_share = total_medicaid / total_adults
  )

# Plot the Medicaid share over time
q3 <- ggplot(medicaid_share_by_year, aes(x = year, y = medicaid_share)) +
  geom_line(color = "black", size = 1.2) +
  geom_point(color = "black", size = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  labs(
    title = "Share of Adult Population with Medicaid Coverage Over Time",
    x = "Year",
    y = "Medicaid Coverage Share (%)"
  ) +
  theme_minimal()

q3

```

4. Plot the share of uninsured over time, separately by states that expanded Medicaid in 2014 versus those that did not. Drop all states that expanded after 2014.

```{r}

# Filter out states that expanded after 2014
medicaid_filtered <- acs_medicaid %>%
  filter(is.na(expand_year) | expand_year <= 2014)

# Create a new grouping variable for expansion in 2014, replacing NA with "Not Expanded"
medicaid_grouped <- medicaid_filtered %>%
  mutate(
    expanded_2014 = if_else(expand_year == 2014, "Expanded in 2014", "Not Expanded"),
    expanded_2014 = replace_na(expanded_2014, "Not Expanded")
  )

# Calculate uninsured share by year and expansion group
uninsured_share_by_group <- medicaid_grouped %>%
  group_by(year, expanded_2014) %>%
  summarise(
    total_uninsured = sum(uninsured, na.rm = TRUE),
    total_adults = sum(adult_pop, na.rm = TRUE),
    uninsured_share = total_uninsured / total_adults,
    .groups = "drop"
  )

# Plot with custom colors (not gray for "Not Expanded")
q4 <- ggplot(uninsured_share_by_group, aes(x = year, y = uninsured_share, 
                                            color = expanded_2014, 
                                            linetype = expanded_2014)) +
  geom_line(size = 1.2) +
  geom_point(size = 2, color = "black") +
  scale_color_manual(
    values = c("Expanded in 2014" = "black", "Not Expanded" = "black"),
    guide = "none"  # hides redundant color legend
  ) +
  scale_linetype_manual(
    values = c("Expanded in 2014" = "solid", "Not Expanded" = "dotted"),
    name = "Medicaid Expansion"
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  theme_minimal()

q4

```


5. Calculate the average percent of uninsured individuals in 2012 and 2015, separately for expansion and non-expansion states. Present your results in a basic 2x2 DD table.
```{r}

# Filter for just 2012 and 2015 and exclude post-2014 expanders
dd_data <- acs_medicaid %>%
  filter(year %in% c(2012, 2015)) %>%
  filter(is.na(expand_year) | expand_year <= 2014) %>%
  mutate(
    expanded_2014 = if_else(expand_year == 2014, "Expanded", "Not Expanded")
  )

# Calculate average uninsured percent for each group
dd_table <- dd_data %>%
  group_by(expanded_2014, year) %>%
  summarise(
    uninsured_rate = sum(uninsured, na.rm = TRUE) / sum(adult_pop, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = year, values_from = uninsured_rate) %>%
  mutate(
    diff = `2015` - `2012`
  )

# Show DD table
q5 <- dd_table

```
6. Estimate the effect of Medicaid expansion on the uninsurance rate using a standard DD regression estimator, again focusing only on states that expanded in 2014 versus those that never expanded.
```{r}
reg.dat <-acs_medicaid  %>% filter(expand_year==2014 | is.na(expand_year), !is.na(expand_ever)) %>%
  mutate(perc_unins=uninsured/adult_pop,
         post = (year>=2014), 
         treat=post*expand_ever)

dd.ins.reg <- lm(perc_unins ~ post + expand_ever + post*expand_ever, data=reg.dat)

# Clustered standard errors by 'State'
cluster_se <- vcovCL(dd.ins.reg, cluster = dd.ins.reg$model$State)

# Get a coeftest with clustered standard errors
clustered_summary <- coeftest(dd.ins.reg, vcov = cluster_se)

# Convert to a tidy format and omit the intercept
tidy_summary <- tidy(clustered_summary) %>%
  filter(term != "(Intercept)") %>%
  select(term, estimate, std.error, statistic, p.value)

# Display the summary using kable
q6 <- kable(tidy_summary, digits = 3, caption = "Regression Output with Clustered SE")

q6
```
7. Include state and year fixed effects in your estimates. Try using the lfe or fixest package to estimate this instead of directly including the fixed effects.
```{r}

reg.dat <- acs_medicaid %>% filter(expand_year==2014 | is.na(expand_year), !is.na(expand_ever)) %>%
  mutate(perc_unins=uninsured/adult_pop,
         post = (year>=2014), 
         treat=post*expand_ever)
m.dd <- lm(perc_unins ~ post + expand_ever + treat, data=reg.dat)
m.twfe <- feols(perc_unins ~ treat | State + year, data=reg.dat)

# 1. Clustered standard errors by 'State' for both models
cluster_se_dd <- vcovCL(m.dd, cluster = m.dd$model$State)
cluster_se_twfe <- vcovCL(m.twfe, cluster = m.twfe$model$State)

# 2. Get coeftest results with clustered standard errors
summary_dd <- coeftest(m.dd, vcov = cluster_se_dd)
summary_twfe <- coeftest(m.twfe, vcov = cluster_se_twfe)

# 3. Tidy the results from both models
tidy_dd <- tidy(summary_dd) %>% filter(term != "(Intercept)") %>% 
  mutate(model = "DD")

tidy_twfe <- tidy(summary_twfe) %>% filter(term != "(Intercept)") %>% 
  mutate(model = "TWFE")

# 4. Combine the results into one data frame
combined_summary <- bind_rows(tidy_dd, tidy_twfe)

# 5. Reshape the data to get the desired format (term + statistic ~ model)
long_form <- combined_summary %>%
  pivot_longer(cols = c(estimate, std.error, statistic, p.value),
               names_to = "statistic", values_to = "value") %>%
  select(term, statistic, model, value)

# 6. Pivot the data to a wide format
final_table <- long_form %>%
  pivot_wider(names_from = model, values_from = value)

# 7. Display the table with kable
q7 <- kable(final_table, digits = 3, caption = "Regression Results (DD and TWFE models) with Clustered SE")

```
8. Repeat the analysis in question 7 but include all states (even those that expanded after 2014). Are your results different? If so, why?
```{r}
# Assuming acs_medicaid dataset is already loaded
# Adjust the dataset to include all states, even those that expanded after 2014
reg.dat <- acs_medicaid %>%
  filter(!is.na(expand_ever)) %>%
  mutate(perc_unins = uninsured / adult_pop,
         post = (year >= 2014),
         treat = post * expand_ever)

# 1. Model using lm() for DD with fixed effects by year and state
m.dd <- lm(perc_unins ~ post + expand_ever + treat + factor(State) + factor(year), data = reg.dat)

# 2. Model using fixest (or lfe) for TWFE
m.twfe <- feols(perc_unins ~ treat | State + year, data = reg.dat)

# 3. Clustered standard errors by 'State' for both models
cluster_se_dd <- vcovCL(m.dd, cluster = reg.dat$State)
cluster_se_twfe <- vcovCL(m.twfe, cluster = reg.dat$State)

# 4. Get coeftest results with clustered standard errors
summary_dd <- coeftest(m.dd, vcov = cluster_se_dd)
summary_twfe <- coeftest(m.twfe, vcov = cluster_se_twfe)

# 5. Tidy the results from both models
tidy_dd <- tidy(summary_dd) %>% 
  filter(term != "(Intercept)") %>%
  mutate(model = "DD")

tidy_twfe <- tidy(summary_twfe) %>% 
  filter(term != "(Intercept)") %>%
  mutate(model = "TWFE")

# 6. Combine the results into one data frame
combined_summary <- bind_rows(tidy_dd, tidy_twfe)

# 7. Reshape the data to get the desired format (term + statistic ~ model)
long_form <- combined_summary %>%
  pivot_longer(cols = c(estimate, std.error, statistic, p.value),
               names_to = "statistic", values_to = "value") %>%
  select(term, statistic, model, value)

# 8. Pivot the data to a wide format
final_table <- long_form %>%
  pivot_wider(names_from = model, values_from = value)

# 9. Display the table with kable
q8<- kable(final_table, digits = 3, caption = "Regression Results (DD and TWFE models) with Clustered SE, Including All States")

```
9. Provide an “event study” graph showing the effects of Medicaid expansion in each year. Use the specification that includes state and year fixed effects, limited to states that expanded in 2014 or never expanded.
```{r}

reg.dat <-acs_medicaid %>% 
  filter(expand_year==2014 | is.na(expand_year), !is.na(expand_ever)) %>%
  mutate(perc_unins=uninsured/adult_pop,
         post = (year>=2014), 
         treat=post*expand_ever)

mod.twfe <- feols(perc_unins~i(year, expand_ever, ref=2013) | State + year,
                  cluster=~State,
                  data=reg.dat)

q9 <- iplot(mod.twfe, 
      xlab = 'Time to treatment',
      main = 'Event study')
```

10. Repeat part 9 but again include states that expanded after 2014. Note: this is tricky…you need to put all states onto “event time” to create this graph.
```{r}

# Prepare the dataset to include all states (including those that expanded after 2014)
reg.dat <- acs_medicaid %>%
  filter(!is.na(expand_ever)) %>%
  mutate(
    perc_unins = uninsured / adult_pop,
    # Create 'post' variable indicating years after 2014
    post = (year >= 2014),
    # Define 'treat' as interaction of 'post' and 'expand_ever'
    treat = post * expand_ever,
    # Create event time variable (relative to year of Medicaid expansion)
    event_time = year - ifelse(expand_ever == 1, expand_year, 2013)  # 2013 is reference year for non-expanders
  )

# Now, add the reference year back to event_time to get actual calendar years
reg.dat1 <- reg.dat %>%
  mutate(time_to_treat = ifelse(expand_ever == TRUE, year-expand_year,-1),
         time_to_treat = ifeslse(time_to_treat <=-4,-4, time_to_treat))  # Adjust event_time to actual calendar year

# 1. Estimate the event study model (including all states)
# Treat 'event_time' as a factor to define the reference category
mod.twfe <- feols(perc_unins ~ i(factor(event_time), expand_ever) | State + year, 
                  cluster = ~State, data = reg.dat)

# 2. Plot the event study graph with the actual 'calendar_year' on the x-axis
q10 <- iplot(mod.twfe, 
      xlab = 'Year',  # Change x-axis label to 'Year'
      main = 'Event Study with States Expanding After 2014',
      xvar = "calendar_year")  # Set xvar to 'calendar_year' for the x-axis

```

```{r}
dir.create("/Users/sushmitarajan/econ470spring2025/Homework5/submission1/results", recursive = TRUE)
save.image("/Users/sushmitarajan/econ470spring2025/Homework5/submission1/results/Hwk5_workspace.RData")
```

