---
title: "analysis"
output: html_document
date: "2025-04-06"
---
```{r setup, include=FALSE}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(scales,tidyverse, ggplot2, dplyr, lubridate, stringr, readxl, data.table, gdata,modelsummary,rdrobust,knitr,broom)

final_data <- readRDS('/Users/sushmitarajan/econ470spring2025/Homework4/data/output/final_ma_data.rds')

```
Summarize the data
Remove all SNPs, 800-series plans, and prescription drug only plans (i.e., plans that do not offer Part C benefits). Provide a box and whisker plot showing the distribution of plan counts by county over time. Do you think that the number of plans is sufficient, too few, or too many?
```{r}

#Filter data for relevant years and valid partc_score
final_data_clean <- final_data %>%
  filter(year %in% 2010:2015, !is.na(partc_score))

#Exclude unwanted plans
# - Remove SNPs
# - Remove 800-series plans (typically employer/union-only)
# - Remove standalone drug plans (PDPs)
final_data_clean <- final_data_clean %>%
  filter(
    snp == "No",
    !(planid >= 800 & planid < 900),
    !(partd == "Y" & plan_type == "PDP")
  )

#Count number of plans per county per year
county_plan_counts <- final_data_clean %>%
  group_by(year, state, county) %>%
  summarise(plan_count = n(), .groups = "drop")

#Create a boxplot to show distribution of plan counts over time
question1 <- ggplot(county_plan_counts, aes(x = factor(year), y = plan_count)) +
  geom_boxplot(fill = "steelblue", color = "navy") +
  coord_cartesian(ylim = c(0, 70)) +
  labs(
    title = "Distribution of Medicare Plan Counts by County (2010–2015)",
    x = "Year",
    y = "Number of Plans per County"
  ) +
  theme_minimal()

# Step 5: Display the plot
  print(question1)

```


Provide bar graphs showing the distribution of star ratings in 2010, 2012, and 2015. How has this distribution changed over time?
```{r}
# Drop NA values before processing
star_dist <- final_data %>%
  filter(year %in% c(2010, 2012, 2015)) %>%
  drop_na(year, Star_Rating) %>%  # Drop rows where year or Star_Rating is NA
  group_by(year, Star_Rating) %>%
  summarise(count = n(), .groups = "drop")

# Create the plot and separate by year
question2 <- ggplot(star_dist, aes(x = as.factor(Star_Rating), y = count, fill = as.factor(year))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Star Rating", y = "Count of Plans", fill = "Year") +
  theme_minimal() +
  facet_wrap(~ year, scales = "free_y")  # Separate by year with independent y-axis scales

# Print the plot
print(question2)

```

Plot the average benchmark payment over time from 2010 through 2015. How much has the average benchmark payment risen over the years?
```{r}

#Calculate the average benchmark payment
avg_benchmark <- final_data %>%
  filter(year >= 2010 & year <= 2015) %>%
  group_by(year) %>%
  summarise(average_benchmark = mean(ma_rate, na.rm = TRUE))  

#Plot the average benchmark payment over time (2010-2015)
question3 <-ggplot(avg_benchmark, aes(x = year, y = average_benchmark)) +
  geom_line(color = "steelblue", size = 1) +   
  geom_point(color = "navy", size = 3) +   
  labs(title = "Average Benchmark Payment Over Time (2010-2015)",
       x = "Year",
       y = "Average Benchmark Payment") +
coord_cartesian(ylim = c(600, 900)) +                  # Expand y-axis range without clipping
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels


print(question3)

```

Plot the average share of Medicare Advantage (relative to all Medicare eligibles) over time from 2010 through 2015. Has Medicare Advantage increased or decreased in popularity? How does this share correlate with benchmark payments?
```{r}
#Calculate the Share of Medicare Advantage Enrollment relative to Total Medicare Eligibles
medicare_share <- final_data %>%
  filter(year >= 2010 & year <= 2015) %>%
  filter(avg_eligibles > 0, avg_enrolled >= 0) %>%
  group_by(year) %>%
  mutate(ma_share = avg_enrolled / avg_eligibles)  # Share of Medicare Advantage

#Calculate the Average Share per Year
avg_share_per_year <- medicare_share %>%
  group_by(year) %>%
  summarise(average_ma_share = mean(ma_share, na.rm = TRUE))  # Average share for each year

#Plot the Trend of Medicare Advantage Share Over Time (2010-2015)
question4<-ggplot(avg_share_per_year, aes(x = year, y = average_ma_share)) +
  geom_line(color = "steelblue", size = 1) +    # Line plot to show the trend
  geom_point(color = "navy", size = 3) +    
  labs(title = "Average Share of Medicare Advantage Over Time (2010-2015)",
       x = "Year",
       y = "Average Share of Medicare Advantage") +   
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + # Rotate x-axis
  scale_y_continuous(labels = label_percent(scale = 100)) 

question4


# Step 4: Check the Correlation Between the Share of Medicare Advantage and Benchmark Payments
# Correlation between Medicare Advantage Share and Benchmark Payments
medicare_share_with_benchmark <- medicare_share %>%
  inner_join(avg_benchmark, by = "year")
correlation <- cor(medicare_share_with_benchmark$ma_share,
                   medicare_share_with_benchmark$average_benchmark,
                   use = "complete.obs")
correlation


```

Estimate ATEs
For the rest of the assignment, we’ll use a regression discontinuity design to estimate the average treatment effect from receiving a marginally higher rating. We’ll focus only on 2010.

Calculate the running variable underlying the star rating. Provide a table showing the number of plans that are rounded up into a 3-star, 3.5-star, 4-star, 4.5-star, and 5-star rating.
```{r}
data_2010 <- final_data %>%
             filter(!is.na(avg_enrollment) & year==2010 & !is.na(partc_score)) %>%
  distinct(contractid, planid, county, .keep_all = TRUE)

#Calculate the raw rating and assign it to a new column
raw_2010 <- data_2010 %>%
  mutate(raw_rating = rowMeans(
    cbind(
      breastcancer_screen, rectalcancer_screen, cv_diab_cholscreen, glaucoma_test, 
      monitoring, flu_vaccine, pn_vaccine, physical_health, mental_health, osteo_test, 
      physical_monitor, primaryaccess, osteo_manage, diab_healthy, bloodpressure, ra_manage, 
      copd_test, bladder, falling, nodelays, doctor_communicate, carequickly, customer_service, 
      overallrating_care, overallrating_plan, complaints_plan, appeals_timely, appeals_review, 
      leave_plan, audit_problems, hold_times, info_accuracy, ttyt_available), 
    na.rm = TRUE)) %>%
  select(contractid, planid, fips, avg_enrollment, state, county, raw_rating, partc_score,
         avg_eligibles, avg_enrolled, premium_partc, risk_ab, Star_Rating,
         bid, avg_ffscost, ma_rate, plan_type) %>% 
    mutate(mkt_share = avg_enrollment/avg_eligibles, 
          HMO=str_detect(plan_type, "HMO"))

#Round star ratings based on raw scores and calculate rounded counts
ratings_2010_summary <- raw_2010 %>%
  filter(Star_Rating %in% c(3, 3.5, 4, 4.5, 5)) %>%  # Focus on specific star ratings
  mutate(
    round_to_3 = as.integer(raw_rating >= 2.75 & raw_rating < 3.00 & Star_Rating == 3),
    round_to_35 = as.integer(raw_rating >= 3.25 & raw_rating < 3.50 & Star_Rating == 3.5),
    round_to_4 = as.integer(raw_rating >= 3.75 & raw_rating < 4.00 & Star_Rating == 4),
    round_to_45 = as.integer(raw_rating >= 4.25 & raw_rating < 4.50 & Star_Rating == 4.5),
    round_to_5 = as.integer(raw_rating >= 4.75 & raw_rating < 5.00 & Star_Rating == 5)
  ) %>%
  group_by(Star_Rating) %>%
  summarise(
    count_round_3 = sum(round_to_3),
    count_round_35 = sum(round_to_35),
    count_round_4 = sum(round_to_4),
    count_round_45 = sum(round_to_45),
    count_round_5 = sum(round_to_5),
    .groups = "drop"
  ) %>%
  mutate(
    total_rounded_up = count_round_3 + count_round_35 + count_round_4 + count_round_45 + count_round_5
  ) %>%
  select(Star_Rating, total_rounded_up)

# Display the results in a table format
kable(ratings_2010_summary, caption = "Summary of Rounded Ratings by Star Rating")



```

Using the RD estimator with a bandwidth of 0.125, provide an estimate of the effect of receiving a 3-star versus a 2.5 star rating on enrollments. Repeat the exercise to estimate the effects at 3.5 stars, and summarize your results in a table.
```{r}
# Load necessary packages
library(kableExtra)
library(dplyr)

# Estimation for the effect of receiving a 3-star vs 2.5-star rating
rd25 <- lm(mkt_share ~ treatment + score, 
           data = raw_2010 %>% 
             filter(raw_rating >= (2.75 - 0.125), 
                    raw_rating <= (2.75 + 0.125), 
                    Star_Rating %in% c(2.5, 3.0)) %>%
             mutate(treatment = (Star_Rating == 3.0), 
                    score = raw_rating - 2.75))

# Estimation for the effect of receiving a 3.5-star vs 3-star rating
rd30 <- lm(mkt_share ~ treatment + score, 
           data = raw_2010 %>% 
             filter(raw_rating >= (3.25 - 0.125), 
                    raw_rating <= (3.25 + 0.125), 
                    Star_Rating %in% c(3.0, 3.5)) %>%
             mutate(treatment = (Star_Rating == 3.5), 
                    score = raw_rating - 3.25))

# Estimation for the effect of receiving a 4.0-star vs 3-star rating
rd35 <- lm(mkt_share ~ treatment + score, 
           data = raw_2010 %>% 
             filter(raw_rating >= (3.75 - 0.125), 
                    raw_rating <= (3.75 + 0.125), 
                    Star_Rating %in% c(3.5, 4.0)) %>%
             mutate(treatment = (Star_Rating == 4.0), 
                    score = raw_rating - 3.75))

# Combine the models into a list
rd <- list(rd25, rd30, rd35)
names(rd) <- c("2.5–3", "3–3.5", "3.5–4")

# Extract coefficient summaries from each regression model
coef_df <- bind_rows(lapply(rd, function(model) {
  coef_summary <- summary(model)$coefficients
  data.frame(
    Model = rep(names(model), nrow(coef_summary)),
    Coefficient = rownames(coef_summary),
    Estimate = coef_summary[, "Estimate"],
    Std_Error = coef_summary[, "Std. Error"],
    t_value = coef_summary[, "t value"],
    p_value = coef_summary[, "Pr(>|t|)"]
  )
}))

# Rename the coefficients based on your labels
coef_df <- coef_df %>%
  dplyr::mutate(
    Coefficient = case_when(
      Coefficient == "treatmentTRUE" ~ "Rounded",
      Coefficient == "score" ~ "Running Score",
      TRUE ~ Coefficient
    )
  )

# Create a kable table with custom column names
coef_df %>%
  kable("latex", caption = "Effect of Star Rating on Enrollment Near Thresholds", 
        col.names = c("Model", "Variable", "Estimate", "Standard Error", "t-value", "p-value"))


```

Repeat your results for bandwidhts of 0.1, 0.12, 0.13, 0.14, and 0.15 (again for 3 and 3.5 stars). Show all of the results in a graph. How sensitive are your findings to the choice of bandwidth?
```{r} 

bandwidths <- c(0.10,0.11, 0.12, 0.13, 0.14, 0.15)

# Create an empty list
results <- list()

# Loop over each bandwidth and estimate the effects for all regressions
for (b in bandwidths) {
  df30 <- lm(mkt_share ~ treatment + score, 
               data = raw_2010 %>%
                 filter(raw_rating >= (2.75 - b),
                        raw_rating <= (2.75 + b),
                        Star_Rating %in% c(2.5, 3.0)) %>%
                 mutate(treatment = (Star_Rating == 3), 
                        score = raw_rating - 2.75))  
  df35 <- lm(mkt_share ~ treatment + score,
           data = raw_2010 %>%
             filter(raw_rating >= (3.25 - b),
                    raw_rating <= (3.25 + b),
                    Star_Rating %in% c(3.0, 3.5)) %>%
             mutate(treatment = (Star_Rating == 3.5),
                    score = raw_rating - 3.25))
  df40 <- lm(mkt_share ~ treatment + score,
           data = raw_2010 %>%
             filter(raw_rating >= (3.75 - b),
                    raw_rating <= (3.75 + b),
                    Star_Rating %in% c(3.5, 4.0)) %>%
             mutate(treatment = (Star_Rating == 4.0),
                    score = raw_rating - 3.75))
  
  # Extract the coefficients (estimates) and standard errors for the effect from each model
  df30_estimate <- tidy(df30)$estimate[2]
  df35_estimate <- tidy(df35)$estimate[2]
  df40_estimate <- tidy(df40)$estimate[2]
  df30_se <- tidy(df30)$std.error[2]
  df35_se <- tidy(df35)$std.error[2]
  df40_se <- tidy(df40)$std.error[2]
  
  # Store the results
  results[[as.character(b)]] <- tibble(
    Bandwidth = b,
    Comparison = c("2.5 to 3 Star", "3 to 3.5 Star", "3.5 to 4 Star"),
    Estimate = c(df30_estimate, df35_estimate, df40_estimate),
    StdError = c(df30_se, df35_se, df40_se)
  )}

# Combine all the results into a single dataframe
results_df <- bind_rows(results)

results_df <- results_df %>%
  mutate(Bandwidth_offset = case_when(
    Comparison == "3.5 to 4 Star" ~ Bandwidth + 0.002,
    TRUE ~ Bandwidth
  ))


# Plot the results
q7 <- ggplot(results_df, aes(x = Bandwidth_offset, y = Estimate, shape = Comparison)) +
  geom_point(size = 4, color = "steelblue", fill = "steelblue", stroke = 1.5) +
  geom_errorbar(aes(ymin = Estimate - StdError, ymax = Estimate + StdError),
                width = 0.001, color = "navy", size = 1) +
  labs(title = "RD Estimates by Star Ratings and Bandwith ",
       x = "Bandwidth",
       y = "Estimated Effect on Market Share",
       shape = "Comparison Type") +
  scale_shape_manual(values = c(16, 17, 15), labels = c("2.5 to 3 Star", "3 to 3.5 Star", "3.5 to 4 Star")) +
  theme_minimal(base_family = "Times") +
  theme(
    legend.position = "top",
    legend.title = element_text(face = "plain", size = 12),
    legend.text = element_text(face = "plain", size = 10),
    axis.title = element_text(face = "plain", size = 12),
    axis.text = element_text(face = "plain", size = 10),
    plot.title = element_text(face = "plain", size = 14, hjust = 0.5))

q7

```

Examine (graphically) whether contracts appear to manipulate the running variable. In other words, look at the distribution of the running variable before and after the relevent threshold values. What do you find?

```{r}
# Create a density plot for 'raw_rating' with vertical lines for each rounding threshold
q8 <- ggplot(raw_2010, aes(x = raw_rating)) +
  geom_density(fill = "grey", alpha = 0.5) +
  geom_vline(aes(xintercept = 2.75, color = "3.0 Star Threshold"),
             linetype = "dashed", size = 0.8, show.legend = TRUE) +
  geom_vline(aes(xintercept = 3.25, color = "3.5 Star Threshold"),
             linetype = "dashed", size = 0.8, show.legend = TRUE) +
  scale_color_manual(
    name = "Rounding Thresholds",
    values = c("3.0 Star Threshold" = "steelblue", "3.5 Star Threshold" = "navy")
  ) +
  labs(
    title = "Distribution of Raw Ratings with Rounding Thresholds",
    x = "Raw Rating", y = "Density"
  ) +
  theme_minimal(base_family = "Times") +
  theme(
    legend.position = "top",
    legend.title = element_text(face = "plain", size = 12),
    legend.text = element_text(size = 10)
  )


q8
```

Similar to question 4, examine whether plans just above the threshold values have different characteristics than contracts just below the threshold values. Use HMO and Part D status as your plan characteristics.
```{r}
# Subset the data for plans just above and below the threshold values
plans_above_below_threshold <- data_2010 %>%
  filter((Star_Rating >= 2.9 & Star_Rating < 3.1) |
    (Star_Rating >= 3.4 & Star_Rating < 3.6))

# Summarize plan characteristics for HMO and Part D status
plan_characteristics <- plans_above_below_threshold %>%
  group_by(Star_Rating) %>%
  summarize(avg_HMO = mean(grepl("HMO", plan_type), na.rm = TRUE),
    avg_part_d = mean(partd == "Yes", na.rm = TRUE),
    count = n())

plan_characteristics_clean <- plan_characteristics %>%
  rename(
    `Star Rating` = Star_Rating,
    `Percent HMO Plans` = avg_HMO,
    `Percent with Part D` = avg_part_d,
    `Number of Plans` = count) %>%
  mutate(
    `Percent HMO Plans` = percent(`Percent HMO Plans`, accuracy = 1),
    `Percent with Part D` = percent(`Percent with Part D`, accuracy = 1))

kable(plan_characteristics_clean, caption = "Plan Characteristics Around Star Rating Thresholds")

```



Summarize your findings from 5-9. What is the effect of increasing a star rating on enrollments? Briefly explain your results.


dir.create("/Users/sushmitarajan/econ470spring2025/Homework4/submission3/results", recursive = TRUE)
save.image("/Users/sushmitarajan/econ470spring2025/Homework4/submission3/results/Hwk4_workspace.RData")
