---
title: "analysis_3"
---
```{r}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, ggplot2, dplyr, lubridate, stringr, readxl, data.table, gdata,modelsummary,rdrobust,readr,lmtest,sandwich,broom,knitr,fixest,scales)

acs_medicaid <- read_tsv('/Users/sushmitarajan/econ470spring2025/Homework5/data/output/acs_medicaid.txt')

acs_medicaid <- acs_medicaid %>%
  mutate(perc_private = (ins_employer + ins_direct)/adult_pop,
         perc_public = (ins_medicare + ins_medicaid)/adult_pop,
         perc_ins = (adult_pop - uninsured)/adult_pop,
         perc_unins = uninsured/adult_pop,
         perc_employer = ins_employer/adult_pop,
         perc_medicaid = ins_medicaid/adult_pop,
         perc_medicare = ins_medicare/adult_pop,
         perc_direct = ins_direct/adult_pop) %>%
  filter(! State %in% c("Puerto Rico", "District of Columbia"))
```

1. Plot the share of the adult population with direct purchase health insurance over time.

```{r}
# Calculate the mean share by year
acs_medicaid_summary <- acs_medicaid %>%
  group_by(year) %>%
  summarize(mean_perc_direct = mean(perc_direct, na.rm = TRUE))

#Plot
q1 <- ggplot(acs_medicaid_summary, aes(x = year, y = mean_perc_direct)) +
  geom_line(color = "black", size = 1.2) +
  geom_point(color = "black", size = 2) +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  labs(
    title = "Share of Adult Population with Direct Purchase Health Insurance",
    x = "Year",
    y = "Direct Purchase Share"
  ) +
  geom_vline(xintercept = 2013.5, color = "red") +
  theme_minimal()

q1
```

2. Discuss the reduction in direct purchase health insurance in later years. Can you list a couple of policies that might have affected the success of the direct purchase insurance market?

```{r}






```

3. Plot the share of the adult population with Medicaid over time.

```{r}
# # Summarize by year: total Medicaid-insured and adult population
medicaid_share_by_year <- acs_medicaid %>%
  group_by(year) %>%
  summarize(mean_perc_medicaid = mean(perc_medicaid, na.rm = TRUE))


# Plot the Medicaid share over time
q3 <- ggplot(medicaid_share_by_year, aes(x = year, y = mean_perc_medicaid )) +
  geom_line(color = "black", size = 1.2) +
  geom_point(color = "black", size = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  labs(
    title = "Share of Adult Population with Medicaid Coverage Over Time",
    x = "Year",
    y = "Medicaid Coverage Share (%)"
  ) +
  geom_vline(xintercept = 2013.5, color = "red")+
   theme_minimal()

q3

```

4. Plot the share of uninsured over time, separately by states that expanded Medicaid in 2014 versus those that did not. Drop all states that expanded after 2014.

```{r}

# Filter out states that expanded after 2014
medicaid_filtered <- acs_medicaid %>%
  filter(is.na(expand_year) | expand_year <= 2014)

# Create a new grouping variable for expansion in 2014, replacing NA with "Not Expanded"
medicaid_grouped <- medicaid_filtered %>%
  mutate(
    expanded_2014 = if_else(expand_year == 2014, "Expanded in 2014", "Not Expanded"),
    expanded_2014 = replace_na(expanded_2014, "Not Expanded")
  )

# Calculate uninsured share by year and expansion group
uninsured_share_by_group <- medicaid_grouped %>%
  group_by(year, expanded_2014) %>%
  summarise(
    mean_perc_unins = mean(perc_unins, na.rm = TRUE),
    .groups = "drop"
  )

# Plot with custom colors (not gray for "Not Expanded")
q4 <- ggplot(uninsured_share_by_group, aes(x = year, y = mean_perc_unins, 
                                            color = expanded_2014, 
                                            linetype = expanded_2014)) +
  geom_line(size = 1.2) +
  geom_point(size = 2, color = "black") +
  scale_color_manual(
    values = c("Expanded in 2014" = "black", "Not Expanded" = "black"),
    guide = "none"  # hides redundant color legend
  ) +
  labs(
    title = "Share of Uninsured over Time",
    x = "Year",
    y = "Uninsured (%)") + 
  scale_linetype_manual(
    values = c("Expanded in 2014" = "dotted", "Not Expanded" = "solid"),
    name = "Medicaid Expansion") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  geom_vline(xintercept = 2013.5, color = "red")+
  theme_minimal()

q4

```


5. Calculate the average percent of uninsured individuals in 2012 and 2015, separately for expansion and non-expansion states. Present your results in a basic 2x2 DD table.
```{r}

# Filter for just 2012 and 2015 and exclude post-2014 expanders
dd_data <- acs_medicaid %>%
  filter(is.na(expand_year) | expand_year <= 2014) %>%
  filter(year %in% c(2012, 2015)) %>%
  group_by(expand_ever, year) %>%
  summarize(uninsured=mean(perc_unins)
  )

# Calculate average uninsured percent for each group
dd_table <- pivot_wider(dd_data, names_from="year", names_prefix="year", values_from="uninsured") %>%
  ungroup() %>%
  mutate(expand_ever=case_when(
    expand_ever==FALSE ~ 'Non-expansion',
    expand_ever==TRUE ~ 'Expansion')
  ) %>%
  rename(Group=expand_ever,
         Pre=year2012,
         Post=year2015)

# Show DD table
q5 <- dd_table

q5

```
6. Estimate the effect of Medicaid expansion on the uninsurance rate using a standard DD regression estimator, again focusing only on states that expanded in 2014 versus those that never expanded.
```{r}
reg_data <-acs_medicaid  %>% filter(expand_year==2014 | is.na(expand_year), !is.na(expand_ever)) %>%
    mutate(post=(year>=2014),
    treat=post*expand_ever)

q6 <- lm(perc_unins~post + expand_ever + treat, data=reg_data)

q6
```
7. Include state and year fixed effects in your estimates. Try using the lfe or fixest package to estimate this instead of directly including the fixed effects.
```{r}

q7 <- feols(perc_unins ~ treat | State + year, data=reg_data)

summary(q7)

```
8. Repeat the analysis in question 7 but include all states (even those that expanded after 2014). Are your results different? If so, why?
```{r}
# Assuming acs_medicaid dataset is already loaded
# Adjust the dataset to include all states, even those that expanded after 2014
reg_data_all <- acs_medicaid %>%
  mutate(treat=case_when(
    year>=expand_year & !is.na(expand_year) ~ 1,
    is.na(expand_year) ~ 0,
    year<expand_year & !is.na(expand_year) ~ 0)
)

# 2. Model using fixest (or lfe) for TWFE
q8_fe <- feols(perc_unins~treat | State + year, data=reg_data_all)

q8_fe


```
9. Provide an “event study” graph showing the effects of Medicaid expansion in each year. Use the specification that includes state and year fixed effects, limited to states that expanded in 2014 or never expanded.
```{r}
# Create the fixed effect  on states that expanded in 2014 based on data from above
mod.twfe1 <- feols(perc_unins~i(year, expand_ever, ref=2013) | State + year,
                  cluster=~State,
                  data=reg_data)
#Plot
q9 <- iplot(mod.twfe1, 
      xlab = 'Time to treatment',
      main = 'Event study with Common Treatment Time')

```

10. Repeat part 9 but again include states that expanded after 2014. Note: this is tricky…you need to put all states onto “event time” to create this graph.
```{r}

# Adding the reference year back to event_time to get actual calendar years
reg_data_all <- reg_data_all %>%
  mutate(time_to_treat=ifelse(expand_ever==TRUE, year-expand_year, -1),
         time_to_treat=ifelse(time_to_treat<=-4, -4, time_to_treat))

# 1. Estimate the event study model (including all states)
# Treat 'event_time' as a factor to define the reference category
mod.twfe2 <- feols(perc_unins~i(time_to_treat, expand_ever, ref=-1) | State + year,
                  cluster=~State,
                  data=reg_data_all)
# 2. Plot the event study graph with the actual 'calendar_year' on the x-axis
q10 <- iplot(mod.twfe2, 
      xlab = 'Year',  # Change x-axis label to 'Year'
      main = 'Event Study with States Expanding After 2014',
      xvar = "calendar_year")  # Set xvar to 'calendar_year' for the x-axis

q10

```

```{r}
dir.create("/Users/sushmitarajan/econ470spring2025/Homework5/submission3/results", recursive = TRUE)
save.image("/Users/sushmitarajan/econ470spring2025/Homework5/submission3/results/Hwk5_workspace.RData")
```

