---
title: "Homework 5"
subtitle: "ECON 470, Spring 2025"
author: "Sushmita Rajan"
format:
  pdf:
    output-file: "rajan_hwk5_s3"
    output-ext:  "pdf"
    header-includes:
      - \usepackage{float}
      - \floatplacement{table}{H}
---

```{r}
#| echo: false
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, ggplot2, dplyr, lubridate, stringr, readxl, data.table, gdata,modelsummary,rdrobust,readr,lmtest,sandwich,broom,knitr,fixest,kableExtra)

load("/Users/sushmitarajan/econ470spring2025/Homework5/submission3/results/Hwk5_workspace.RData") 

```

\[Click here to visit my GitHub repository\](https://github.com/sarajan03/econ470spring2025/tree/main/Homework5)

\newpage 

\noindent 1. Plot the share of insured individuals with direct purchase health insurance over time.

```{r}
#| echo: false
#| label: fig-directplot
#| fig-cap: "Share of Individuals with Direct Purchase"

print(q1)

```

\newpage

\noindent 2. Discuss the reduction in direct purchase health insurance in later years. Can you list a couple of policies that might have affected the success of the direct purchase insurance market?

> In the later years, we observe a reduction in the share of adults with direct purchase health insurance. This decline can be attributed to a few significant factors. One of the most important policies influencing this trend was the Affordable Care Act (ACA), which provided more affordable options through the Medicaid expansion and the health insurance marketplaces. Another key policy was the introduction of the individual mandate penalty, which incentivized individuals to seek other coverage options beyond direct purchase.

\newpage

\noindent 3.Plot the share of the adult population with Medicaid over time.

```{r}
#| echo: false
#| label: fig-medicaidplot
#| fig-cap: "Share of Individuals with Medicaid"

print(q3)

```

\newpage

\noindent 4. Plot the share of uninsured over time, separately by states that expanded Medicaid in 2014 versus those that did not. Drop all states that expanded after 2014.

```{r}
#| echo: false
#| label: fig-uninsuranceplot
#| fig-cap: "Average Uninsured by Medicaid Expansion"

print(q4)

```

\newpage

\noindent  5. Calculate the average percent of uninsured individuals in 2012 and 2015, separately for expansion and non-expansion states. Present your results in a basic 2x2 DD table.

```{r}
#| echo: false
#| label: tbl-ddtable
#| tbl-cap: "DD Table for Medicaid Expansion"

knitr::kable(q5, 
             col.names=c("Group", "Pre", "Post"),
             digits=2,
             booktabs = TRUE,
             escape=F,
             align="l",
             format='latex') %>% 
  kable_styling(full_width=F)
```

\newpage

\noindent 6. Estimate the effect of Medicaid expansion on the uninsurance rate using a standard DD regression estimator, again focusing only on states that expanded in 2014 versus those that never expanded.

```{r}
#| echo: false
#| label: tbl-ddregs
#| tbl-cap: "DD Estimates for Medicaid Expansion"

modelsummary(q6,
             coef_rename=c("postTRUE" = "Post 2014","expand_everTRUE"="Expand",
                           "treat" = "Post x Expand"),
             gof_omit='DF|F|Lik|AIC|BIC|Adj')


```

\newpage

\noindent 7. Include state and year *fixed effects* in your estimates. Try using the `lfe` or `fixest` package to estimate this instead of directly including the fixed effects.

```{r}
#| echo: false
#| label: tbl-ddregs2
#| tbl-cap: "DD Estimates for Medicaid Expansion with TWFE"

models <- list("Standard DD" = q6, 
               "TWFE" = q7)
modelsummary(models,
             coef_rename=c("postTRUE" = "Post 2014","expand_everTRUE"="Expand",
                           "treat" = "Post x Expand"),
             gof_omit='DF|F|Lik|AIC|BIC|Adj')
```

\newpage

\noindent 8. Repeat the analysis in question 7 but include all states (even those that expanded after 2014). Are your results different? If so, why?

```{r}
#| echo: false
#| label: tbl-ddregs3
#| tbl-cap: "DD Estimates for Medicaid Expansion with Staggered Treatment"

models <- list("Standard DD" = q6, 
               "TWFE" = q7,
               "Time-varying Treatment" = q8_fe)
modelsummary(models,
             coef_rename=c("postTRUE" = "Post 2014","expand_everTRUE"="Expand",
                           "treat" = "Post x Expand"),
             gof_omit='DF|F|Lik|AIC|BIC|Adj')
```

\newpage

\noindent 9. Provide an \"event study\" graph showing the effects of Medicaid expansion in each year. Use the specification that includes state and year fixed effects, limited to states that expanded in 2014 or never expanded.

```{r}
#| echo: false
#| label: fig-eventstudy1
#| fig-cap: "Event Study with Common Treatment Time"
q9 <- iplot(mod.twfe1, 
      xlab = 'Time to treatment',
      main = 'Event study with Common Treatment Time')
```

\newpage

\noindent 10. Repeat part 9 but again include states that expanded after 2014. Note: this is tricky...you need to put all states onto \"event time\" to create this graph.

```{r}
#| echo: false
#| label: fig-eventstudy2
#| fig-cap: "Event Study with Staggered Treatment"

q10 <- iplot(mod.twfe2, 
      xlab = 'Year',  # Change x-axis label to 'Year'
      main = 'Event Study with States Expanding After 2014',
      xvar = "calendar_year")  # Set xvar to 'calendar_year' for the x-axis

q10
```
